Leveraging open platforms to manage and distribute streams of versioned data

### 2017-07-11 ms.tex
more editing of the text

Almost all useful databases are living--they \emph{should} get better
through time either by the addition of new data or by improving the
quality of existing data. For example a dataset on plant characteristics
might become obsolete as taxonomic changes occur. Updating the dataset might only
require a small amount of work, but if there is no
way to distribute the updated dataset, then that work has to be re-done by
every user of that dataset. Moreover, those required changes add steps
between the canonical (or master) dataset and the new analysis (and publication),
making reproducibility of the new analysis more difficult than it needs
to be. If the canonical dataset is versioned--and versions are
accessible--the up-to-date data can be distributed, used, and analyses
based on any of the released versions data can be reproduced.

@@

Providing versioned data is laudable goal, but implementation has been
difficult. The static model of data distribution widely used in science, where
datasets are archived either as supplementary materials for a journal article
or under its own DOI, is simply not designed to provide streams of versioned
data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
model is that data releases are mostly triggered by publication of academic
papers. But for a living dataset, there may be multiple releases made in
between papers. This is particularly the case with the discovery of small
errors. For any medium-sized dataset that is put to reuse, errors will
inevitably be found, and there needs to be some easy  way to correct them and
distribute the updated dataset.

To overcome the limitations of static datasets, many research groups have
setup dedicated webservers to host and distribute data. In these systems,
datasets areas maintained in a structure database, such as SQL, and a front
end to this database is written, often using a platform such as Ruby on rails.
When designed well, such systems are very useful for updating or adding to a
dataset. However, the infrastructure investment (designing maintaining, and
hosting the database) is clearly overkill for most projects and is a very
substantial barrier to entry. And the system requires ongoing funds to keep it
going, and possibly maintenance. Moreover, the existence of a live webserver,
does not in itself guarantee ongoing access to different versions of a
dataset.

@@

In brief, the workflow we present borrows best-practices for software
development  and applies them to the challenge of maintaining and distributing
versioned data. Software developers maintain a core set of code which produces
a binary executable file that can be installed on a users local  computers. In
our system, data developers maintain a core set of files, which produce an
organised dataset that can be similarly be ``installed'' on a users local
computer. In either the development of software or data, successive versions
are released over time (called ``releases''). The similarity in workflow
between sofwtare and data then allows us to deploy the same technological
platforms that are used in software development, for the development and
distribution of data. Importantly, the system we describe uses well-
established and tools, which ensures high-level performance and stability.

@@

\texttt{Git} is a version control system that tracks line-by-line changes in text
files. Git is used extensively in software development, but is increasingly
also being applied to scientific code and data management \citep{Ram-2013,
Perkel-2016}. Some key features of git that make attractive for management of data
include:
\begin{itemize}
  \item Tracks all changes in monitored files. These changes are visible to
  anyone interacting with the repository.
  \item Allows users to annotate changes (`commits') with informative messages
  detailing the changes.
  \item Is a distributed system, meaning anyone who downloads the repository
  can access the full history and contribute changes.
  \item Can easily handle file sizes of at least 50Mb and repositories up to 1Gb.
\end{itemize}

\subsection{Github platform for storing git repositories}

The website Github (\smurl{www.github.com}) is a commercial platform for
hosting and interacting with git repositories. In practice, each dataset
should be a separate git repository on github (see Table \ref{tab:examples} for
examples). Although mainly used for
computer code, Github is now also being use to manage scientific data
\citep{Perkel-2016}. Github has a double role in our workflow:
\begin{enumerate}
  \item It provides a platform for data contributors to sync their files
  and correspond about changes in the dataset.
  \item It hosts a stream of data releases, thus acting as a central
  point for distribution of the data.
\end{enumerate}

@@

\begin{itemize}
  \item \textbf{Issues}: Each repository has tab where contributors can document issues in the data. These issue tags then provide a centralised point for describing a problem and discussing potential solutions.
  \item \textbf{Pull requests}: Much of the activity on Github is on opens-ource projects with many contributors. An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
  \item \textbf{Releases}: Github is designed to host releases (in this case compiled dataset), associated with a particular commit to the repository.
  \item \textbf{Tags}: Git commits and releases can be tagged with an informative label.
  \item \textbf{It's free}: All open-source projects are hosted for free on Github.
  \item \textbf{Quality}: The Github servers are designed for managing enetrpise level projects, and as such can deliver multiple requests for releases and have sophisticated capabilities.
  \item \textbf{DOIs}: Github repos can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
\end{itemize}

@@

To address the issue of how successive releases should be labelled, we suggest
applying the theory of ``semantic versioning'' developed for software
distribution. This system uses (the potentially familiar) notation of form
``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted
(REFS), labelling of data versions presents a similar problem to the labelling
of software releases: signalling to users the types of change that has
occurred between successive versions. Applying semantic versioning to data, a
change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error
correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement,
for example adding a new study to a meta-analysis dataset (while otherwise
maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies
a very major revision, for example improving the entire structure of the
dataset and adding new columns.

\subsection{\texttt{R}, \texttt{CRAN}, \texttt{datastorr}}

While Github can host a stream of versioned data releases, we also provide a
way for users to retrieve the data. One method is to point and click using a
web browser, then download the data from the Github releases tab. But
computational users will want something more sophisticated, allowing automated
retrieval of different versions. Github makes this possible by providing a
thorough API from which releases can be accessed programatically. And ideally
load it directly into their analysis platform. As many of the users of small-
to-medium sized datasets use the open source language \texttt{R}, we also
prioritized easy of use from within \texttt{R}. Moreover, the  Comprehensive R
Archive Network (CRAN, \smurl{cran.r-project.org}) offers a facility for users
to write their own packages, which are then archived and able to be installed
by anyone running R.

To distribute/retrieve versioned data releases from Github, we have developed
a  work flow that enables dataset maintainers to write a customised R package
for their dataset. This workflow uses another R package called
\texttt{datastorr} to handle all the interactions with github. For example, the package
\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).

@@

\subsection{Make your own data package}

Table \ref{tab:examples} gives details for XXX example datasets already using
the distribution system described here. Each of these is a versioned dataset,
designed for a particular research goal, and also explicitly flexible such
that they can grow in the future. In all three cases, the user needs to
install a few R packages. The data versions themselves are stored as Github
Releases (as well as if desired on DataDryad or FigShare), then any of the
versions of the data can be downloaded and used with only one of \texttt{R}
code.

>>>>

Almost all useful databases are living--they \emph{should} get better
through time either by the addition of new data or by improving the
quality of existing data. For example a dataset on plant characteristics
might become outdated as taxonomic changes occur. Updating the dataset might only
require a small amount of work, but if there is no
way to distribute the updated dataset, then that work has to be re-done by
every user of that dataset. Moreover, those required changes add steps
between the canonical (or master) dataset and the new analysis (and publication),
making reproducibility of the new analysis more difficult than it needs
to be. If the canonical dataset is versioned--and versions are
accessible--the up-to-date data can be distributed, used, and analyses
based on any of the released versions data can be reproduced.

@@

While the benefits of versioned data are fairly clear, determining how best to implement it has been anything but. The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
or under its own DOI, is simply not designed to provide streams of versioned
data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
model is that data releases are mostly triggered by publication of academic
papers. But for a living dataset, there may be multiple releases made in
between papers. This is particularly the case with the discovery of small
errors. For any medium-sized dataset that is put to reuse, errors will
inevitably be found, and there needs to be some easy  way to correct them and
distribute the updated dataset.

To overcome the limitations of static datasets, many research groups have
setup dedicated webservers to host and distribute data. In these systems,
datasets areas maintained in a structure database, such as SQL, and a front
end to this database is written, often using a platform such as Ruby on rails.
When designed well, such systems are very useful for updating or adding to a
dataset. However, the infrastructure investment (designing maintaining, and
hosting the database) is clearly overkill for most projects and is a very
substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many groups. Moreover, the existence of a live webserver,
does not in itself guarantee ongoing access to different versions of a
dataset.

@@

In brief, the workflow we present, which we call Lightweight Versioned Data (LVD), borrows best-practices for software
development and applies them to the challenge of maintaining and distributing
versioned data. Software developers maintain a core set of code which produces
a binary executable file that can be installed on a users local  computers. With a LVD system, data developers maintain a core set of files, which produce an
organised dataset that can be similarly be ``installed'' on a user's local
computer. In either the development of software or data, successive versions
are released over time (called ``releases''). The similarity in workflow
between sofwtare and data then allows us to deploy the same technological
platforms that are used in software development, for the development and
distribution of data. Importantly, LVD uses well-
established and tools, which ensures high-level performance and stability.

@@

\texttt{Git} is a version control system that tracks line-by-line changes in text
files. Git is used extensively in software development, but is increasingly
also being applied to scientific code and data management \citep{Ram-2013,
Perkel-2016}. Git is attractive for data management because it: 
\begin{itemize}
  \item Tracks all changes in monitored files. These changes are visible to
  anyone interacting with the repository.
  \item Allows users to annotate changes (`commits') with informative messages
  detailing the changes.
  \item Is a distributed system, meaning anyone who downloads the repository
  can access the full history and contribute changes.
  \item Can easily handle file sizes of at least 50Mb and repositories up to 1Gb.
\end{itemize}

\subsection{GitHub platform for storing git repositories}

The website GitHub (\smurl{www.github.com}) is a commercial platform for
hosting and interacting with git repositories. In practice, each dataset
should be a separate git repository on github (see Table \ref{tab:examples} for
examples). Although mainly used for
computer code, Github is now also being use to manage scientific data
\citep{Perkel-2016}. Github has a double role in our workflow:
\begin{enumerate}
  \item It provides a platform for data contributors to sync their files
  and correspond about changes in the dataset.
  \item It hosts a stream of data releases, thus acting as a central
  point for distribution of the data.
\end{enumerate}

@@

\begin{itemize}
  \item \textbf{Issues}: Each repository has tab where contributors can document issues in the data. These issue tags then provide a centralised point for describing a problem and discussing potential solutions.
  \item \textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors. An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
  \item \textbf{Releases}: GitHub is designed to host releases (in this case compiled datasets), associated with a particular commit to the repository.
  \item \textbf{Tags}: Git commits and releases can be tagged with an informative label.
  \item \textbf{It's free}: All open-source projects are hosted for free on GitHub.
  \item \textbf{Quality}: The GitHub servers are designed for managing enetrpise level projects, and as such can deliver multiple requests for releases and have sophisticated capabilities.
  \item \textbf{DOIs}: Github repos can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
\end{itemize}

@@

To address the issue of how successive releases should be labelled, we suggest
applying the theory of ``semantic versioning'' developed for software
distribution. This system uses (the potentially familiar) notation of form
``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted
(REFS), labelling of data versions presents a similar problem to the labelling
of software releases: signalling to users the types of change that has
occurred between successive versions. Applying semantic versioning to data, a
change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error
correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement,
for example adding a new study to a meta-analysis dataset (while otherwise[]
maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies
a very major revision, for example improving the entire structure of the
dataset and adding new columns.

\subsection{\texttt{datastorr} and dataset-specific R packages}

While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform. Our solution is for researchers releasing a dataset to write a customized R package, specific to their dataset. For example, the package
\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples). To enable this, we have developed a new R package called \texttt{datastorr}, which does all the heavy lifting both in terms of creating templates for the data-specific R packages and for interacting with the GitHub API. With \texttt{datastorr}, setting up a new data-specific package is straightforward and requires minimal computational skills.

@@

We have created \texttt{datastorr} to make LVD easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}). However, we can imagine alternative methods for interacting with GitHub and building templates; so while useful, adopting LVD does not critically depend on using \texttt{datastorr}. Furthermore, the concept of LVD is general and we would love to see implementations in \texttt{Python} or other languages.
